Supplement: Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 supplement.tex
simplified derivation of parameter estimator

We use an EM algorithm to estimate the continuous parameters of the DSL, e.g. $\theta$.
Suppressing dependencies on $\mathcal{D}$, the EM updates are
\begin{align}
\label{maximizeStep}  \theta& = \argmax_\theta \log P(\theta) + \sum_x \expect_{Q_x}\left[\log \probability\left[p|\theta \right] \right]\\
  Q_x(p)&\propto \probability[x|p]\probability[p|\theta]
  \end{align}
In the M step of EM we will update $\theta$ by instead maximizing a lower bound on $\log \probability[p|\theta]$.
We write $c(e,p)$ to mean the number of times that primitive $e$ was used in program $p$; $c(p)= \sum_{e\in \mathcal{D}}c(e,p)$ to mean the total number of primitives used in program $p$; $R(p)$ to mean the sequence of types input to sample in Alg.~1 of the main paper. Jensen's inequality gives an intuitive lower bound on the likelihood of a program $p$:
\begin{align*}
  \log \probability[p|\theta]&\stackrel{+}{ = }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - \sum_{\tau\in R(p)} \log \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  &\stackrel{+}{\geq }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{\tau\in R(p)} \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  & = \sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{e\in \mathcal{D}} r(e,p)\theta_e
\end{align*}
where we have defined
$$r(e:\tau',p)\triangleq\sum_{\tau\in R(p)} \indicator[\text{canUnify}(\tau,\tau')]$$

Returning to equation~\ref{maximizeStep}:
\begin{align*}
\log P(\theta) +& \sum_x \expect_{Q_x}\left[\log \probability\left[p|\theta \right] \right]\geq \\
\log   \text{Dir}(\theta|\alpha) +& \sum_x\sum_{e\in \mathcal{D}}\expect_{Q_x}\left[ c(e,p)\right]\log \theta_e -\\
&\sum_x  \expect_{Q_x}\left[c(p)\log \sum_{e\in \mathcal{D}}r(e,p)\theta_e \right]
  \end{align*}

Differentiate with respect to $\theta_e$:
\begin{align}
  & \frac{\alpha + \sum_x\expect_{Q_x}\left[ c(e,p)\right]}{\theta_e} -
\theta_e\sum_x  \expect_{Q_x}\left[c(p)\frac{r(p,e)}{\sum_{e'\in \mathcal{D}} r(e',p)\theta_{e'}} \right]  
  \end{align}
and set to zero:
%\begin{align*}
%  \frac{c(e, p)}{\theta_e} &= c(p)\frac{r(e, p)}{\sum_{e'\in\mathcal{D}} r(e', p)\theta_{e'}}
%\end{align*}
%or more succinctly,
\begin{align}
  \frac{c(x,p)}{\theta_x} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
where $N=c(p)$ and $a(x) = r(x, p)$. From here forward, we similarly let
$c(x)=c(x,p)$ for notational convenience.

@@

Now what we are actually after is the parameters that maximize the joint log probability of the data and parameters $\theta$, which we  write $J$:
\begin{align}
  J& = L + \log \text{Dir}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
We identify the parameters of the prior over $\theta$, $\alpha$, as  pseudocounts that are added to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$).

>>>>

%% We use an EM algorithm to estimate the continuous parameters of the DSL, e.g. $\theta$.
%% Suppressing dependencies on $\mathcal{D}$, the EM updates are
%% \begin{align}
%% \label{maximizeStep}  \theta& = \argmax_\theta \log P(\theta) + \sum_x \expect_{Q_x}\left[\log \probability\left[p|\theta \right] \right]\\
%%   Q_x(p)&\propto \probability[x|p]\probability[p|\theta]
%%   \end{align}
%% In the M step of EM we will update $\theta$ by instead maximizing a lower bound on $\log \probability[p|\theta]$.
We write $c(e,p)$ to mean the number of times that primitive $e$ was used in program $p$; $c(p)= \sum_{e\in \mathcal{D}}c(e,p)$ to mean the total number of primitives used in program $p$; $R(p)$ to mean the sequence of types input to sample in Alg.~1 of the main paper. Jensen's inequality gives an intuitive lower bound on the likelihood of a program $p$:
\begin{align*}
  \log \probability[p|\theta]&\stackrel{+}{ = }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - \sum_{\tau\in R(p)} \log \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  &\stackrel{+}{\geq }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{\tau\in R(p)} \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  & = \sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{e\in \mathcal{D}} r(e,p)\theta_e
\end{align*}
where we have defined
$$r(e:\tau',p)\triangleq\sum_{\tau\in R(p)} \indicator[\text{canUnify}(\tau,\tau')]$$

%% Returning to equation~\ref{maximizeStep}:
%% \begin{align*}
%% \log P(\theta) +& \sum_x \expect_{Q_x}\left[\log \probability\left[p|\theta \right] \right]\geq \\
%% \log   \text{Dir}(\theta|\alpha) +& \sum_x\sum_{e\in \mathcal{D}}\expect_{Q_x}\left[ c(e,p)\right]\log \theta_e -\\
%% &\sum_x  \expect_{Q_x}\left[c(p)\log \sum_{e\in \mathcal{D}}r(e,p)\theta_e \right]
%%   \end{align*}

Differentiate with respect to $\theta_e$ and set to zero:
%% \begin{align}
%%   & \frac{\alpha + \sum_x\expect_{Q_x}\left[ c(e,p)\right]}{\theta_e} -
%% \theta_e\sum_x  \expect_{Q_x}\left[c(p)\frac{r(p,e)}{\sum_{e'\in \mathcal{D}} r(e',p)\theta_{e'}} \right]  
%%   \end{align}

\begin{align*}
 \frac{c(e, p)}{\theta_e} &= c(p)\frac{r(e, p)}{\sum_{e'\in\mathcal{D}} r(e', p)\theta_{e'}}
\end{align*}
%or more succinctly,
\begin{align}
  \frac{c(x,p)}{\theta_x} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
where $N=c(p)$ and $a(x) = r(x, p)$. From here forward, we similarly let
$c(x)=c(x,p)$ for notational convenience.

@@

Now what we are actually after is the parameters that maximize the joint log probability of the data and parameters $\theta$, which we  write $J$:
\begin{align}
  J& = L + \log \text{Dir}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
We identify the parameters of the prior over $\theta$, $\alpha$, as  pseudocounts that are added to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$),
and so we estimate $\theta_e$ as the ratio of the sum of the number of times that $e$ was used, plus the pseudocounts $\alpha$, with the number of times that it could have been used, $a(e)$.
